A Partitioning Algorithm for Maximum Common (Connected) Subgraphs\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/K503058/1 and EP/M508056/1]}

### 2017-02-10 paper/paper.tex
Editing

\usepackage[ruled,vlined]{algorithm2e}
\usepackage{MnSymbol,wasysym}
\usepackage{latexsym}
\usepackage{algpseudocode}

@@

@@

The most successful existing approaches to solving MCS are reformulation to the
maximum clique problem on one hand, and constraint programming
\cite{DBLP:conf/cp/McCreeshNPS16} on the other. In the clique encoding, an
\textit{association graph} is created with one node for each pair in the set
$\{(u,v) \mid u \in \V(G), v \in \V(H) \}$ and and edges for each pair of
mappings that may be chosen together. A maximum clique in this graph represents
a maximum common subgraph of $G$ and $H$.

In the constraint programming approach, we have a variable corresponding to
each vertex in $G$, and each variable initially has one value in its domain for
each vertex in $H$ to which it may be mapped, plus an additional value $\bot$
signifying that the vertex remains unmapped. A backtracking search is performed
in which values are assigned to variables, and propagation rules are used to
remove values from the domains of unassigned variables.

@@

We introduce a branch and bound algorithm, \McSplit, which uses a new
vertex-labelling scheme to avoid explicitly maintaining the set of feasible
mappings for each vertex in $\V(G)$. As we will explore in
Section~\ref{sec:comparison}, the algorithm may be viewed as a version of the
constraint programming forward-checking (CP-FC) algorithm using a compact data
structure that exploits the nature of the MCS problem.  It is sufficiently
different, however, that it is helpful to describe the algorithm without
reference to previous approaches.

@@

We will use the graphs $G$ and $H$ in Figure~\ref{fig:alg1} to demonstrate the
algorithm.  These graphs have a maximum common subgraph with four vertices; the
mapping $\{1a, 2f, 3d, 5b\}$ (where vertex $1$ is assigned to vertex $a$ and so
on) is one example.

@@

The algorithm proceeds by building up a mapping $M$, starting with the empty
mapping $\{\}$. Select a vertex in $\V(G)$ as the first vertex to
be mapped; we will arbitrarily choose vertex $1$. Each of the vertices in $\V(H)$
to which $1$ may be mapped will be tried in turn, and finally the possibility where
$1$ remains unmatched will be tried.

@@

Next, extend the mapping by pairing a vertex in $G$ with a vertex in $H$ of the
same label; we will choose to map vertex $2$ to vertex $d$, giving $M=\{1a,
2d\}$ (Figure~\ref{fig:alg3}).  Each unmatched vertex $v \in \V(G)$ is labelled
with a two-character bit string, indicating whether $v$ is adjacent to each of
the two matched vertices in $\V(G)$ (vertices $1$ and $2$).  For example, vertex
$3$ is labelled $01$, indicating that it is not adjacent to $1$ but is adjacent
to $2$.  Labels are given to vertices in $\V(H)$ in a similar fashion, showing
whether each vertex is adjacent to $a$ and $d$.

Observe that our invariant is maintained: we can extend $M$ by a vertex pairing if
and only if the two vertices have the same label. Note also that each label is prefixed
by the label at the previous level of the search tree (Figure~\ref{fig:alg2}).

@@

The algorithm backtracks when the incumbent best mapping is at least as large
as a calculated bound given $M$ and the current labelling. To demonstrate how
this bound is calculated, we consider the situation one level deeper in the search
tree shown in Figure~\ref{fig:alg4}. Three vertex labels are used: 100, 101, and 111.
The first two of these only appear in one graph, and therefore vertices with these
labels cannot be added to the mapping. The final label, 111, appears once in $G$ and
twice in $H$, and therefore at most one pair with this label can be added to $M$.
Thus, the upper bound on matching size is $4$. In general, for each label $l$
we add to the bound $\min(|\{ v \in \V(G) : \vtxlabel(v)=l\}|,
|\{ v \in \V(H) : \vtxlabel(v)=l \}|)$.

@@

@@

\paragraph{Efficient representation of label classes} The algorithm as
described so far requires $\BigO{n^2}$ space at each node of the search tree,
since labels may be up to $n_G$ in length. We can reduce this to $\BigO{n}$
by storing a pair $\langle P,T \rangle$ for each label $l$ that is used, where $P$
is the set of vertices in $\V(G)$ labelled $l$, and $T$ is the set of vertices
in $\V(H)$ labelled $l$. Since there are $n_G + n_H$ vertices, at most this number
of label-classes can exist at once, and the total length of the $P$ and $T$ sets
is at most $n_G + n_H$. We can gain further efficiency by discarding during search
any label-class that has an empty $P$ or $T$ set.

@@

\paragraph{Full algorithm} The full algorithm is shown as
Algorithm~\ref{jtAlg}.  The main recursive procedure, $\FuncSty{expand}$, has
two parameters.  The parameter $\AlgVar{future}$ is a list of label classes,
each represented as a $\langle P, T \rangle$ pair as described above.  The
parameter $M$ is the current mapping of vertices.  On each call to
$\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be added to $M$
if and only if $v$ and $w$ belong to the same label class in $M$.

@@

We start our search at the function $\FuncSty{McSplit}$ (\lineref{McSplitFun}),
with graphs $G$ and $H$ as inputs.  This function returns the size of the
largest mapping.  In \lineref{FirstExpandCall} a call is made to
$\FuncSty{expand}$; at this point we have a single label-class containing all
vertices, and the mapping $M$ is empty.

@@

If we have vertex labels, we replace ${\langle V(G),V(H) \rangle}$ in line 28 of
the algorithm with a set containing a label class for each label that appears in
both $\V(G)$ and $\V(H)$. In the remainder of this section, \textit{labels} refers to
edge labels.

For the unlabelled, directed case, let $A_G$ and $A_H$ be the adjacency matrices
of $G$ and $H$ respectively, as stored in memory. For each
vertex pair $(t,u)$ in graph $G$ or $H$, the adjacency matrix entry 
takes the value 0 if the vertices are
not adjacent, 1 if the two vertices share a single edge in the direction $t
\rightarrow u$, 2 if they share a single edge in the direction $u \rightarrow
t$, and 3 if there are edges in both directions. Where lines 14-19 of the
basic algorithm split the pair $\langle P',T' \rangle$ two ways, according to whether
vertices are ajacent to $v$ (resp. $w$), we now perform a four-way split where
each vertex is classified according to the label on its adjacency matrix from $v$
(resp. $w$). This is shown in Algorithm~\ref{labDirAlg}, where $L=\{0,1,2,3\}$.

@@

In the labelled, undirected case, each adjacency-matrix entry contains the edge
label, or a null entry 0 if no edge exists between the pair of vertices.  we
can also use Algorithm~\ref{labDirAlg}, by letting $L$ be the union of $\{0\}$
with the set of all labels that appear in the input graphs. Since there may be
up to $n_G + n_H$ distinct labels, the loop in Algorithm~\ref{labDirAlg} may
execute up to $n_G + n_H$ times, and this variation of the algorithm therefore
requires $\BigO{n^2}$ time per search node.  It is possible to modify the
algorithms to run in $\BigO{n \log n}$ per search node; we will give a brief
outline of the method. First, run lines 17-19 of Algorithm~\ref{jtAlg} to
create a new label-class of vertices that are not adjacent to $v$ or $w$, and
remove these vertices from $\langle P',T' \rangle$. Next, sort $P'$ and $T'$ in
ascending order of the label on the edge from $v$ or $w$ to each vertex. We can
then create the label classes corresponding to each edge label by
simulataneously traversing $P'$ and $T'$ from left to right.

The labelled, directed case is similar to its undirected counterpart, except that
each element $A[u][v]$ in an adjacency matrix is a pair $(l_1, l_2)$, where
$l_1$ is the label on the edge $u \rightarrow v$ (or 0 if no edge exists) and $l_2$
is the label on the reverse edge.

>>>>

\usepackage[ruled,vlined]{algorithm2e}
\usepackage{MnSymbol,wasysym}
\usepackage{latexsym}
\usepackage{algpseudocode}
\usepackage{amsmath}

@@

This paper introduces a new branch and bound algorithm for MCS that improves
state-of-the-art run times on unlabelled, undirected instances by well over an
order of magnitude, and uses only quadratic memory in the order of the input
graphs, allowing us to solve larger instances than previously possible.

@@

The most successful existing approaches to solving MCS are reformulation to the
maximum clique problem on one hand, and constraint programming
\cite{DBLP:conf/cp/McCreeshNPS16} on the other. In the clique encoding, an
association graph is created with one node for each pair in the set $\{(u,v)
\mid u \in \V(G), v \in \V(H) \}$ and an edge for each pair of mappings that
may be chosen together. A maximum clique in this graph corresponds directly to
a maximum common subgraph of $G$ and $H$.

In the constraint programming approach, we have a variable corresponding to
each vertex in $G$, and each variable initially has one value in its domain for
each vertex in $H$ to which it may be mapped, plus an additional value $\bot$
signifying that the vertex remains unmapped. A backtracking search is performed
in which values are assigned to variables, and propagation rules are used to
remove infeasible values from domains.

@@

This paper's main contribution is a branch and bound algorithm, \McSplit,
which uses a new vertex-labelling scheme to avoid explicitly maintaining the
set of feasible mappings for each vertex in $\V(G)$. As we will explore in
Section~\ref{sec:comparison}, the algorithm may be viewed as a version of the
constraint programming forward-checking (CP-FC) algorithm using a compact data
structure that exploits the nature of the MCS problem.  It is sufficiently
different, however, that it is helpful to describe the algorithm without
reference to previous approaches.

@@

We will use the graphs $G$ and $H$ in Figure~\ref{fig:alg1} to demonstrate the
algorithm.  These graphs have a maximum common subgraph with four vertices; one
example is the mapping $\{1a, 2f, 3d, 5b\}$ (where vertex $1$ is assigned to
vertex $a$, $2$ is assigned to $f$, and so on).

@@

The algorithm builds up a mapping $M$, starting with the empty mapping
$\emptyset$. Select a vertex in $\V(G)$ as the first vertex to be mapped; in
our example we will arbitrarily choose vertex $1$. Each of the vertices in
$\V(H)$ to which $1$ may be mapped will be tried in turn, and finally the
possibility where $1$ remains unmatched will be tried.

@@

Next, extend the mapping by pairing a vertex in $G$ with a vertex in $H$ of the
same label; we will choose to map vertex $2$ to vertex $d$, giving $M=\{1a,
2d\}$ (Figure~\ref{fig:alg3}).  Each unmapped vertex $v \in \V(G)$ is labelled
with a two-character bit string, indicating whether $v$ is adjacent to each of
the two mapped vertices in $\V(G)$ (vertices $1$ and $2$).  For example, vertex
$3$ is labelled $01$, indicating that it is not adjacent to $1$ but is adjacent
to $2$.  Labels are given to unmapped vertices in $\V(H)$ in a similar fashion,
showing whether each vertex is adjacent to $a$ and $d$.

Our invariant is maintained: we can extend $M$ by a vertex pairing if
and only if the two vertices have the same label. Note also that each label is prefixed
by the label at the previous level of the search tree (Figure~\ref{fig:alg2}).

@@

The algorithm backtracks when the incumbent best mapping is at least as large
as a calculated bound given $M$ and the current labelling. To demonstrate how
this bound is calculated, we consider the situation one level deeper in the
search tree shown in Figure~\ref{fig:alg4}.

@@

Three vertex labels are used: 100,
101, and 111.  The first two of these only appear in one graph, and therefore
there is no way to add a pair of vertices with label 100 or 101 to the mapping.
The final label, 111, appears once in $G$ and twice in $H$, and therefore at
most one pair with this label can be added to $M$.  Thus, the upper bound on
matching size is $|M| + 1 = 4$. The general formula for the upper bound is
as follows, where $L$ is the set of labels used in both graphs.

\begin{multline*}
    \mathit{bound} = |M| + \sum_{l \in L} \min\big(|\{ v \in \V(G) : \vtxlabel(v)=l\}|, \\
        |\{ v \in \V(H) : \vtxlabel(v)=l \}|\big)
\end{multline*}

@@

\paragraph{Efficient representation of label classes} The algorithm as
described so far requires $\BigO{n^2}$ space at each level of the search tree,
since labels may be up to $n_G$ bits in length. We can reduce the memory
requirement to $\BigO{n}$ per level by storing a pair $\langle P,T \rangle$ for
each label $l$ that is used, where $P$ is the set of vertices in $\V(G)$
labelled $l$, and $T$ is the set of vertices in $\V(H)$ labelled $l$. Since
there are $n_G + n_H$ vertices in the two graphs, at most $n_G + n_H$
label-classes can exist at once, and the total length of the $P$ and $T$ sets
is at most $n_G + n_H$.  We can gain further efficiency by discarding during
search any label-class that has an empty $P$ or $T$ set.

@@

\paragraph{Full algorithm} The full algorithm is shown as
Algorithm~\ref{jtAlg}.  The recursive procedure, $\FuncSty{expand}$, has
two parameters.  The parameter $\AlgVar{future}$ is a list of label classes,
each represented as a $\langle P, T \rangle$ pair as described above.  The
parameter $M$ is the current mapping of vertices.  On each call to
$\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be added to $M$
if and only if $v$ and $w$ belong to the same label class in $M$.

@@

We start our search at the function $\FuncSty{McSplit}$ (\lineref{McSplitFun}),
with graphs $G$ and $H$ as inputs.  This function returns the size of the
largest mapping.  In \lineref{FirstExpandCall} the initial call is made to
$\FuncSty{expand}$; at this point we have a single label-class containing all
vertices, and the mapping $M$ is empty.

@@

\paragraph{Vertex labels} If we have vertex labels, we replace ${\langle
V(G),V(H) \rangle}$ in \lineref{FirstExpandCall} of the algorithm with a set
containing a label class for each label that appears in both $\V(G)$ and
$\V(H)$. Loops on vertices can be handled without difficulty; assuming without
loss of generality that all vertex labels are positive integers, we can negate
the labels of all vertices that have loops.

The remainder of this section considers edges that are directed or labelled.

\paragraph{Directed, without edge labels} Let $A_G$ and $A_H$ be the adjacency matrices
of $G$ and $H$ respectively, as stored in memory. For each
vertex pair $(t,u)$ in graph $G$ or $H$, the adjacency matrix entry 
takes the value 0 if the vertices are
not adjacent, 1 if the two vertices share a single edge in the direction $t
\rightarrow u$, 2 if they share a single edge in the direction $u \rightarrow
t$, and 3 if there are edges in both directions. Where lines 14-19 of the
basic algorithm split the pair $\langle P',T' \rangle$ two ways, according to whether
vertices are ajacent to $v$ (resp. $w$), we now perform a four-way split where
each vertex is classified according to the label on its adjacency matrix from $v$
(resp. $w$). This is shown in Algorithm~\ref{labDirAlg}, where $L=\{0,1,2,3\}$.

@@

\paragraph{Undirected, with edge labels} In this case, each adjacency-matrix
entry contains the edge label, or a null entry 0 if no edge exists between the
pair of vertices.  we can also use Algorithm~\ref{labDirAlg}, by letting $L$ be
the union of $\{0\}$ with the set of all labels that appear in the input
graphs. Since there may be up to $n_G + n_H$ distinct labels, the loop in
Algorithm~\ref{labDirAlg} may execute up to $n_G + n_H$ times, and this
variation of the algorithm therefore requires $\BigO{n^2}$ time per search
node.  It is possible to modify the algorithms to run in $\BigO{n \log n}$ per
search node; we will give a brief outline of the method. First, run lines 17-19
of Algorithm~\ref{jtAlg} to create a new label-class of vertices that are not
adjacent to $v$ or $w$, and remove these vertices from $\langle P',T' \rangle$.
Next, sort $P'$ and $T'$ in ascending order of the label on the edge from $v$
or $w$ to each vertex. We can then create the label classes corresponding to
each edge label by simulataneously traversing $P'$ and $T'$ from left to right.

\paragraph{Directed, with edge labels} This case is similar to its undirected
counterpart, except that each element $A[u][v]$ in an adjacency matrix is a
pair $(l_1, l_2)$, where $l_1$ is the label on the edge $u \rightarrow v$ (or 0
if no edge exists) and $l_2$ is the label on the reverse edge.
